\textbf{The Structure of Chaos: An Empirical Comparison of Fractal Physiology Complexity Indices using NeuroKit2}

### 2022-07-19 analysis.Rmd
update

@@

Leveraging this tool, the goal of this study is to empirically compare a large number of complexity indices, inspect how they relate to one another, and derive recommendations for indices selection. More specifically, we will quantify the complexity of various types of signals with varying degrees of noise using 115 of the predominantly used indices that are available for computation using *NeuroKit2*. Note that, even though it is one of the largest comparison to date to our knowledge (and covers the most commonly used metrics), this list is by no means exhaustive: new indices are being developed, such as for instance *symmetropy* [@girault2022]. We will then project the results on a latent space through factor analysis, and review the various indices that we find the most relevant and interesting in regards to their representation of the latent dimensions. This analysis will be complemented by hierarchical clustering.

@@

```{r message=FALSE, warning=FALSE, include=FALSE, results='hide'}
cache <- FALSE
```
```{r signals, message=FALSE, warning=FALSE, fig.height=16, fig.width=10, echo=FALSE, fig.cap = "Different types of simulated signals, to which was added 5 types of noise (violet, blue, white, pink, and brown) with different intensities. For each signal type, the first row shows the signal with a minimal amount of noise, and the last with a maximal amount of noise.", cache=FALSE}
library(tidyverse)
library(easystats)
library(patchwork)

@@

The combination of these parameters resulted in a total of `r 128*6*6*5` signal iterations. For each of them, we computed `r length(unique(df$Index))` complexity indices, and additional basic metrics such as the *length* of the signal and its dominant *frequency*. We also included a *random* number to make sure that our our dimensionality analyses accurately discriminate this unrelated feature. The parameters used (such as the time-delay $\tau$ or the embedding dimension) are documented in the data generation script. For a complete description of the various indices included, please refer to NeuroKit's documentation at **https://neuropsychology.github.io/NeuroKit**, as well as to the data generation script.

@@

```{r loadings, message=FALSE, warning=FALSE, fig.width=12, fig.height=18, fig.cap="Factor loadings of the complexity indices, colored by the factor they represent the most (center). On the left, the median computation times and on the right, the archetypicity (the inverse of factor profile complexity, i.e., the extent to which each index is a pure representative of its dominant factor. This measure is low for indices that equally load on different factors)."}
n <- 13
rez <- parameters::factor_analysis(data, cor = r, n = n, rotation = "varimax", sort = TRUE, fm = "ml")
# rez <- parameters::principal_components(data, n = 15, sort = TRUE)
# rez

@@

```{r varexplained, echo=FALSE, fig.cap="Variance of the whole dataset of indices explained by the subselection. Each line represents a random number of selected variables. The green line represents the optimal order (i.e., the relative importance) that maximizes the variance explained. The dotted blue line represents the cumulative relative median computation time of the selected indices, and shows that FuzzyMSEn and MFDFA indices are the most costly algorithms.", fig.height=7, fig.width=10, message=FALSE, warning=FALSE, cache=cache}
dat <- select(data, -Random, -Frequency, -Noise, -Length)

@@

p5 <- varex |>
  mutate(Order = as.factor(Order)) |>
  ggplot(aes(x = Order, y = VarExplained)) +
  geom_line(aes(group = Iteration), color = "black", alpha = 0.05, size = 0.3) +
  geom_line(aes(group = 1), data = time, color = "#1E88E5", alpha = 0.5, size = 1, linetype = "dotted") +
  geom_line(data = optim, size = 1, color = "#388E3C") +
  annotate("text",
    x = 1, y = 0.9,
    label = paste0("Max. R² = ", insight::format_value(max(optim$VarEx, na.rm = TRUE), as_percent = TRUE)),
    color = "black", hjust = 0, vjust = 0) +
  annotate("text", label=labels$label, x=labels$Order, y=labels$VarExplained + 0.01,
           angle=45, hjust = 0, vjust = 0, color="#388E3C") +
  labs(x = "Number of Selected Indices", y = "Variance Explaiend", title = "Explanatory Power of Indices Subset") +
  scale_y_continuous(labels = scales::percent, expand=c(0, 0)) +
  coord_cartesian(ylim = c(0.5, 0.95)) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
# ggsave("figures/varexplained-1.png", p5, width=7, height=5)
p5
```

>>>>

```{r, echo = FALSE, warning=FALSE, message=FALSE}
options(digits = 3)

knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 450,
  fig.path = "./figures/"
)
```

@@

Leveraging this tool, the goal of this study is to empirically compare a large number of complexity indices, inspect how they relate to one another, and derive recommendations for indices selection. More specifically, we will quantify the complexity of various types of signals with varying degrees of noise using 112 of the predominantly used indices that are available for computation using *NeuroKit2*. Note that, even though it is one of the largest comparison to date to our knowledge (and covers the most commonly used metrics), this list is by no means exhaustive: new indices are being developed, such as for instance *symmetropy* [@girault2022]. We will then project the results on a latent space through factor analysis, and review the various indices that we find the most relevant and interesting in regards to their representation of the latent dimensions. This analysis will be complemented by hierarchical clustering.

@@

```{r message=FALSE, warning=FALSE, include=FALSE, results='hide'}
cache <- TRUE
```
```{r signals, message=FALSE, warning=FALSE, fig.height=16, fig.width=10, echo=FALSE, fig.cap = "Different types of simulated signals, to which was added 5 types of noise (violet, blue, white, pink, and brown) with different intensities. For each signal type, the first row shows the signal with a minimal amount of noise, and the last with a maximal amount of noise.", cache=FALSE}
library(tidyverse)
library(easystats)
library(patchwork)

@@

The combination of these parameters resulted in a total of `r 128*6*6*5` signal iterations. For each of them, we computed 112 complexity indices, and additional basic metrics such as the *length* of the signal and its dominant *frequency*. We also included a *random* number to make sure that our our dimensionality analyses accurately discriminate this unrelated feature. The parameters used (such as the time-delay $\tau$ or the embedding dimension) are documented in the data generation script. For a complete description of the various indices included, please refer to NeuroKit's documentation at **https://neuropsychology.github.io/NeuroKit**, as well as to the data generation script.

@@

```{r loadings, message=FALSE, warning=FALSE, fig.width=12, fig.height=18, fig.cap="Factor loadings of the complexity indices, colored by the factor they represent the most (center). On the left, the median computation times and on the right, the archetypicity - the inverse of factor profile complexity, i.e., the extent to which each index is a pure representative of its dominant factor. This measure is low for indices that equally load on different factors)."}
n <- 13
rez <- parameters::factor_analysis(data, cor = r, n = n, rotation = "varimax", sort = TRUE, fm = "ml")
# rez <- parameters::principal_components(data, n = 15, sort = TRUE)
# rez

@@

```{r varexplained, echo=FALSE, fig.cap="Variance of the whole dataset of indices explained by the subselection. Each line represents a random number of selected variables. The green line represents the optimal order (i.e., the relative importance) that maximizes the variance explained. The dotted blue line represents the cumulative relative median computation time of the selected indices, and shows that MFDFA and multiscale indices are the most costly algorithms.", fig.height=7, fig.width=10, message=FALSE, warning=FALSE, cache=cache}
dat <- select(data, -Random, -Frequency, -Noise, -Length)

@@

p5 <- varex |>
  mutate(Order = as.factor(Order)) |>
  ggplot(aes(x = Order, y = VarExplained)) +
  geom_line(aes(group = Iteration), color = "black", alpha = 0.05, size = 0.3) +
  geom_line(aes(group = 1), data = time, color = "#1E88E5", alpha = 0.5, size = 1, linetype = "dotted") +
  geom_line(data = optim, size = 1.2, color = "#388E3C") +
  annotate("text",
    x = 1, y = 0.9,
    label = paste0("Max. R² = ", insight::format_value(max(optim$VarEx, na.rm = TRUE), as_percent = TRUE)),
    color = "black", hjust = 0, vjust = 0) +
  annotate("text", label=labels$label, x=labels$Order, y=labels$VarExplained + 0.02,
           angle=45, hjust = 0, vjust = 0, color="#388E3C") +
  labs(x = "Number of Selected Indices", y = "Variance Explaiend", title = "Explanatory Power of Indices Subset") +
  scale_y_continuous(labels = scales::percent, expand=c(0, 0)) +
  coord_cartesian(ylim = c(0.5, 0.95)) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
# ggsave("figures/varexplained-1.png", p5, width=7, height=5)
p5
```

### 2022-07-19 manuscript.Rmd
update

authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com) and Annabel Chen (AnnabelChen@ntu.edu.sg).
abstract: |
  Complexity quantification, through entropy, information theory and fractal dimension indices, is gaining a renewed traction in psychophsyiology, as new measures with promising qualities emerge from the computational and mathematical advances. Unfortunately, few studies compare the relationship and objective performance of the plethora of existing metrics, in turn hindering reproducibility, replicability, consistency, and clarity in the field. Using the NeuroKit2 Python software, we computed a list of 115 complexity indices on signals varying in their characteristics (noise, length and frequency spectrum). We then systematically compared the indices by their computational weight, their representativeness of a multidimensional space of latent dimensions, and empirical proximity with other indices. Based on these considerations, we propose that a selection of 12 indices, together representing 85.97\\% of the total variance of all indices, might offer a parsimonious and complimentary choice in regards to the quantification of the complexity of time series. Our selection includes *CWPEn*, *Line Length (LL)*, *BubbEn*, *MSWPEn*, *MFDFA (Max)*, *Hjorth Complexity*, *SVDEn*, *MFDFA (Width)*, *MFDFA (Mean)*, *MFDFA (Peak)*, *MFDFA (Fluctuation)*, *AttEn*. Elements of consideration for alternative subsets are discussed, and data, analysis scripts and code for the figures are open-source.

>>>>

authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com) and Annabel Chen (AnnabelChen@ntu.edu.sg).
abstract: |
  Complexity quantification, through entropy, information theory and fractal dimension indices, is gaining a renewed traction in psychophsyiology, as new measures with promising qualities emerge from the computational and mathematical advances. Unfortunately, few studies compare the relationship and objective performance of the plethora of existing metrics, in turn hindering reproducibility, replicability, consistency, and clarity in the field. Using the NeuroKit2 Python software, we computed a list of 112 complexity indices on signals varying in their characteristics (noise, length and frequency spectrum). We then systematically compared the indices by their computational weight, their representativeness of a multidimensional space of latent dimensions, and empirical proximity with other indices. Based on these considerations, we propose that a selection of 12 indices, together representing 85.97\\% of the total variance of all indices, might offer a parsimonious and complimentary choice in regards to the quantification of the complexity of time series. Our selection includes *CWPEn*, *Line Length (LL)*, *BubbEn*, *MSWPEn*, *MFDFA (Max)*, *Hjorth Complexity*, *SVDEn*, *MFDFA (Width)*, *MFDFA (Mean)*, *MFDFA (Peak)*, *MFDFA (Fluctuation)*, *AttEn*. Elements of consideration for alternative subsets are discussed, and data, analysis scripts and code for the figures are open-source.
